Stochastic Variance-Reduced Policy Gradient

### 2018-02-08 svrpg.tex
fixed expected values in Lemma 9

%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
Under Assumptions \ref{ass:bounded_score} ans \ref{ass:M2}, the expected squared norm of the true gradient $\gradJ{\vtheta_t^{s+1}}$, for appropriate choices of $\alpha_t\geq0$ and $\beta_t>0$, can be bounded as follows:
\[
	\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1}}{\Psi_t} + \frac{d_tV}{N\Psi_t}
	+\frac{f_tW}{B\Psi_t},
\]
	where
\begin{align*}
	&R_t^{s+1}\coloneqq \Ets{J(\vtheta_t^{s+1}) - c_t\norm[]{\vtheta_t^{s+1}-\wt{\vtheta}^s}^2}, \\
	&c_{m-1} = 0, \\
	&c_t = c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t+\frac{\alpha_t^2L^2}{B}\right)+\frac{\alpha_t^2L^3}{2B}, \\
	&\Psi_t = \alpha_t\left(\frac{1}{2}-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right), \\
	&d_t = \frac{\alpha_t}{2}\left(1+2c_{t+1}+\alpha_tL+2\alpha_tc_{t+1}\right), \\
	&f_t = \alpha_t^2\frac{\Gamma(L+2c_{t+1})}{2},
\end{align*}
where $L=\max\left\{L1,L2\right\}$, \ie the greater of the Lipschitz constants from Lemmas \ref{lemma:lsmooth} and \ref{lemma:gsmooth}.

In particular, the following constraints on $\alpha_t$ and $\beta_t$ are sufficient:
\begin{align*}
&0\leq\alpha_t < \frac{2(1-\nicefrac{2c_{t+1}}{\beta_t})}{L+2c_{t+1}} \\
&\beta_t > 2c_{t+1}.
\end{align*}
\end{restatable}
\begin{proof}
	We have:
	\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \label{eq:11}\\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \label{eq:12}\\
	&\geq
	\Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} -\frac{\alpha_t}{2}\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2}, \label{eq:13}
\end{align}
where~\eqref{eq:11} is from the L-smoothness of $J(\vtheta)$ and \eqref{eq:12} is from the SVRPG update.
{\color{blue} Inequality~\ref{eq:13} follows from Lemma~\ref{lemma:aux0} by noticing that $\nabla J(\vtheta^{s+1}_t)$ is a deterministic function given $\vtheta^{s+1}_t$. As a consequence we can directly apply Lemma~\ref{lemma:aux0} to lower-bound $\EVV[t|s]{\langle \nabla J(\vtheta^{s+1}_t), \blacktriangledown J(\vtheta^{s+1}_t) \rangle}$ by $\norm[2]{\nabla J(\vtheta^{s+1}_t}^2 - f(\vtheta^{s+1}_t, \wt{\vtheta}^s)$.}
\todopir{is \eqref{eq:11} true just for smooth functions? Does it require convexity/concavity? Ref?}

@@

+2\alpha_t\Ets{\left|\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_t^{s+1}-\wt{\vtheta}^s}\right|} \nonumber\\ 

@@

+2\alpha_t\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}\norm[]{\vtheta_t^{s+1}-\wt{\vtheta}^s}} \nonumber\\ 

@@

+2\alpha_t\Ets{\frac{1}{2\beta_t}\norm[]{\gradJ{\vtheta_t^{s+1}}}^2+\frac{\beta_t}{2}\norm[]{\vtheta_t^{s+1}-\wt{\vtheta}^s}^2} \label{eq:16a}\\ 

@@

where~\eqref{eq:14a} is obtained using the triangular inequality,~\eqref{eq:14} is from the SVRPG update,~\eqref{eq:15} is from Lemma~\ref{lemma:aux0},~\eqref{eq:16a} is from Cauchy-Schwarz inequality, and~\eqref{eq:16} is from Young's inequality in the `Peter-Paul' variant.
Let's us consider the following function:

@@

    % &= \Ets{J(\vtheta_{t+1}^{s+1}) - c_{t+1}\norm[]{\vtheta_{t+1}^{s+1}-\tilde{\vtheta}^s}^2} \nonumber\\
	%&\geq	\Ets{J(\vtheta_t^{s+1})+\frac{\alpha_t}{2}\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\

@@

	&\geq \Ets{J(\vtheta_t^{s+1})+\frac{\alpha_t}{2}\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} 

@@

	&\qquad-2c_{t+1}\alpha_t\Ets{\frac{1}{2\beta_t}\norm[]{\gradJ{\vtheta_t^{s+1}}}^2+\frac{\beta_t}{2}\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2} \nonumber\\ 

@@

	&= \Ets{J(\vtheta_t^{s+1})} - c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t\right)\Ets{\norm[]{\vtheta_{t}^{s+1}-\tilde{\vtheta}}^2} \nonumber\\

@@

	+\frac{\alpha_t}{2}\left(1-\frac{2c_{t+1}}{\beta_t}\right)\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \nonumber\\

@@

	&\geq  \Ets{J(\vtheta_t^{s+1})} - c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t\right)\Ets{\norm[]{\vtheta_{t}^{s+1}-\tilde{\vtheta}}^2} \nonumber\\

@@

	+\frac{\alpha_t}{2}\left(1-2\frac{c_{t+1}}{\beta_t}\right)\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \nonumber\\

@@

	& = \Ets{J(\vtheta_t^{s+1}) - \left(c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t+\frac{\alpha_t^2L^2}{B}\right)+\frac{\alpha_t^2L^3}{2B}\right)\norm[]{\vtheta_{t}^{s+1}-\tilde{\vtheta}}^2} \nonumber\\

@@

	+\alpha_t\left(\frac{1}{2}-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right)\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \nonumber\\

@@

	+\Psi_t\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2}

@@

where~\eqref{eq:17} is from~\eqref{eq:13} noticing that $\EVV[t|s]{\norm[]{\nabla J(\vtheta^{s+1}_t)}^2} = \EVV[t|s]{\norm[]{\nabla J(\vtheta^{s+1}_t)}^2}$,~\eqref{eq:18} is from~\eqref{eq:16},~\eqref{eq:19} is from Lemma~\ref{lemma:aux2}, and (\ref{eq:20}) is from Assumption \ref{ass:REINFORCE}.

>>>>

%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
Under Assumptions \ref{ass:bounded_score} ans \ref{ass:M2}, the expected squared norm of the true gradient $\gradJ{\vtheta_t^{s+1}}$, for appropriate choices of $\alpha_t\geq0$ and $\beta_t>0$, can be bounded as follows:
\[
	\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1}}{\Psi_t} + \frac{d_tV}{N\Psi_t}
	+\frac{f_tW}{B\Psi_t},
\]
	where
\begin{align*}
	&R_t^{s+1}\coloneqq \Ets[t-1]{J(\vtheta_t^{s+1}) - c_t\norm[]{\vtheta_t^{s+1}-\wt{\vtheta}^s}^2}, \\
	&c_{m-1} = 0, \\
	&c_t = c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t+\frac{\alpha_t^2L^2}{B}\right)+\frac{\alpha_t^2L^3}{2B}, \\
	&\Psi_t = \alpha_t\left(\frac{1}{2}-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right), \\
	&d_t = \frac{\alpha_t}{2}\left(1+2c_{t+1}+\alpha_tL+2\alpha_tc_{t+1}\right), \\
	&f_t = \alpha_t^2\frac{\Gamma(L+2c_{t+1})}{2},
\end{align*}
where $L=\max\left\{L1,L2\right\}$, \ie the greater of the Lipschitz constants from Lemmas \ref{lemma:lsmooth} and \ref{lemma:gsmooth}.

In particular, the following constraints on $\alpha_t$ and $\beta_t$ are sufficient:
\begin{align*}
&0\leq\alpha_t < \frac{2(1-\nicefrac{2c_{t+1}}{\beta_t})}{L+2c_{t+1}} \\
&\beta_t > 2c_{t+1}.
\end{align*}
\end{restatable}
\begin{proof}
	We have:
	\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \label{eq:11}\\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \label{eq:12}\\
	&\geq
	\Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2N}\Vars{g(\cdot\vert\wt{\vtheta}^s)} -\frac{\alpha_t}{2}\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2}, \label{eq:13}
\end{align}
where~\eqref{eq:11} is from the L-smoothness of $J(\vtheta)$ \cite{nesterov2013introductory} and \eqref{eq:12} is from the SVRPG update.
Inequality~\ref{eq:13} follows from Lemma~\ref{lemma:aux0} by noticing that $\nabla J(\vtheta^{s+1}_t)$ is a deterministic function given $\vtheta^{s+1}_t$. As a consequence, we can directly apply Lemma~\ref{lemma:aux0} with $\vphi(\vtheta_t^{s+1})\coloneqq\gradJ{\vtheta_t^{s+1}}$.

@@

+2\alpha_t\Ets[t-1]{\left|\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_t^{s+1}-\wt{\vtheta}^s}\right|} \nonumber\\ 

@@

+2\alpha_t\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}\norm[]{\vtheta_t^{s+1}-\wt{\vtheta}^s}} \nonumber\\ 

@@

+2\alpha_t\Ets[t-1]{\frac{1}{2\beta_t}\norm[]{\gradJ{\vtheta_t^{s+1}}}^2+\frac{\beta_t}{2}\norm[]{\vtheta_t^{s+1}-\wt{\vtheta}^s}^2} \label{eq:16a}\\ 

@@

where~\eqref{eq:14a} is obtained using the triangular inequality,~\eqref{eq:14} is from the SVRPG update,~\eqref{eq:15} is from Lemma~\ref{lemma:aux0} with $\vphi(\vtheta_t^{s+1})\coloneqq\vtheta_t^{s+1}-\tilde{\vtheta}^s$,~\eqref{eq:16a} is from Cauchy-Schwarz inequality, and~\eqref{eq:16} is from Young's inequality in the `Peter-Paul' variant.
Let us consider the following function:

@@

@@

	+ \Ets[t-1]{\frac{\alpha_t}{2}\norm[]{\gradJ{\vtheta_t^{s+1}}}^2 }

@@

	&\qquad-2c_{t+1}\alpha_t\Ets[t-1]{\frac{1}{2\beta_t}\norm[]{\gradJ{\vtheta_t^{s+1}}}^2+\frac{\beta_t}{2}\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2} \nonumber\\ 

@@

	&= \Ets[t-1]{J(\vtheta_t^{s+1})} - c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t\right)\Ets[t-1]{\norm[]{\vtheta_{t}^{s+1}-\tilde{\vtheta}}^2} \nonumber\\

@@

	+\frac{\alpha_t}{2}\left(1-\frac{2c_{t+1}}{\beta_t}\right)\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \nonumber\\

@@

	&\geq  \Ets[t-1]{J(\vtheta_t^{s+1})} - c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t\right)\Ets[t-1]{\norm[]{\vtheta_{t}^{s+1}-\tilde{\vtheta}}^2} \nonumber\\

@@

	+\frac{\alpha_t}{2}\left(1-2\frac{c_{t+1}}{\beta_t}\right)\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \nonumber\\

@@

	& = \Ets[t-1]{J(\vtheta_t^{s+1}) - \left(c_{t+1}\left(1+\alpha_t\beta_t+\alpha_t+\frac{\alpha_t^2L^2}{B}\right)+\frac{\alpha_t^2L^3}{2B}\right)\norm[]{\vtheta_{t}^{s+1}-\tilde{\vtheta}}^2} \nonumber\\

@@

	+\alpha_t\left(\frac{1}{2}-\frac{c_{t+1}}{\beta_t}-\frac{\alpha_tL}{2}-\alpha_tc_{t+1}\right)\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \nonumber\\

@@

	+\Psi_t\Ets[t-1]{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2}

@@

where~\eqref{eq:17} is from~\eqref{eq:13} noticing that $\EVV[t|s]{\norm[]{\nabla J(\vtheta^{s+1}_t)}^2} = \EVV[t-1|s]{\norm[]{\nabla J(\vtheta^{s+1}_t)}^2}$,~\eqref{eq:18} is from~\eqref{eq:16},~\eqref{eq:19} is from Lemma~\ref{lemma:aux2}, and (\ref{eq:20}) is from Assumption \ref{ass:REINFORCE}.
